---
title: "Step 02 — Clinical EDA"
subtitle: "Read the imbalance, choose the metrics"
params:
  solved: true
format:
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
---

## Where we are

Cumulative snapshot: load (Step 00) and wrangle (Step 01) are given below; the new
beat is exploratory analysis — a stratified summary table and one figure that read
the class imbalance and motivate our metric choice.

<!--
Nota docente (IT): step HANDS-ON (~22 min). load + wrangle sono "given" (beat
precedenti, già risolti agli step 00-01). Il beat nuovo è la EDA: `tbl_summary`
stratificata per outcome + un boxplot di `ejection_fraction` (il predittore
clinico chiave: chi sopravvive ha una frazione d'eiezione più alta). Messaggio
chiave: l'evento (morte) è ~32%, quindi l'accuracy inganna (predire sempre
"survived" dà già ~68%). Da qui scegliamo DUE metriche: ROC-AUC (capacità di
ranking) e PR-AUC (focus sulla classe positiva = morte; baseline = prevalenza
~0.32). Nei prossimi step le mostriamo SEMPRE insieme.
-->

# Given — load & wrangle {#sec-given}

From Steps 00–01.

```{r given}
#| code-fold: true
library(here)
library(rio)
library(tidyverse)
library(janitor)

hf_raw <- import(here("data-raw", "heart_failure.csv"), setclass = "tibble")

hf <- hf_raw |>
  clean_names() |>
  select(-time) |>
  mutate(
    outcome = factor(
      if_else(death_event == 1, "died", "survived"),
      levels = c("died", "survived"),
    ),
  ) |>
  select(-death_event) |>
  mutate(across(c(anaemia, diabetes, high_blood_pressure, sex, smoking), factor))

glimpse(hf)
```

# Explore — summarise & visualise {#sec-eda}

A stratified summary across the outcome, then a look at `ejection_fraction` (the
heart's pumping efficiency) by outcome.

```{r eda}
#| code-fold: false
library(gtsummary)

# Stratified summary table ----
hf |>
  tbl_summary(
    by = outcome,
    include = c(
      age, sex, ejection_fraction, serum_creatinine, serum_sodium,
      creatinine_phosphokinase, platelets, anaemia, diabetes,
      high_blood_pressure, smoking,
    ),
  )
```

```{r ef-plot}
#| code-fold: false
# Ejection fraction by outcome ----
ef_plot <- hf |>
  ggplot(aes(x = outcome, y = ejection_fraction, fill = outcome)) +
  geom_boxplot() +
  theme_minimal()

ggsave(here("output", "ejection_fraction_by_outcome.png"), plot = ef_plot, width = 6, height = 4)
ef_plot
```

## Reading the table

The event (`died`) is the minority class — roughly **32%** of patients. With this
imbalance, plain accuracy is misleading: a model that always predicts "survived"
would already score ~68%. So from the next step on we score every model with
**two** complementary metrics:

- **ROC-AUC** — how well the model *ranks* patients by risk, regardless of
  prevalence. A coin flip scores 0.5.
- **PR-AUC** (precision–recall) — how well it does specifically on the *positive*
  (death) class. Its no-skill baseline is the **prevalence** itself (~0.32), so it
  is far more honest than ROC-AUC when the event is rare.

Watching both at once is the point: they do not always agree, and knowing *why* is
half the skill.
