---
title: "Step 02 — Clinical EDA"
subtitle: "Read the imbalance, choose the metric"
params:
  solved: true
format:
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
---

## Where we are

Cumulative snapshot: load (Step 00) and wrangle (Step 01) are given below; the new
beat is exploratory analysis — a stratified summary table and one figure that read
the class imbalance and motivate our metric choice.

<!--
Nota docente (IT): step HANDS-ON (~22 min). load + wrangle sono "given" (beat
precedenti, già risolti agli step 00-01). Il beat nuovo è la EDA: `tbl_summary`
stratificata per outcome + un boxplot di `age`. Messaggio chiave: l'evento è raro
(~13%), quindi l'accuracy inganna -> sceglieremo ROC-AUC come metrica primaria.
-->

# Given — load & wrangle {#sec-given}

From Steps 00–01.

```{r given}
#| code-fold: true
library(here)
library(rio)
library(tidyverse)
library(janitor)

indo_raw <- import(here("data-raw", "indo_rct.csv"), setclass = "tibble")

indo <- indo_raw |>
  clean_names() |>
  select(outcome, age, risk, gender, sod, rx, type) |>
  filter(!is.na(age)) |>
  mutate(outcome = factor(outcome, levels = c("0_no", "1_yes")))

glimpse(indo)
```

# Explore — summarise & visualise {#sec-eda}

A stratified summary across the outcome, then a quick look at `age` by outcome.

```{r eda}
#| code-fold: false
library(gtsummary)

# Stratified summary table ----
indo |>
  tbl_summary(
    by = outcome,
    include = c(age, risk, gender, sod, rx, type),
  )
```

```{r age-plot}
#| code-fold: false
# Age distribution by outcome ----
age_plot <- indo |>
  ggplot(aes(x = outcome, y = age, fill = outcome)) +
  geom_boxplot() +
  theme_minimal()

ggsave(here("output", "age_by_outcome.png"), plot = age_plot, width = 6, height = 4)
age_plot
```

## Reading the table

The event (`1_yes`) is the minority class — roughly **13%** of patients. With such
imbalance, plain accuracy is misleading: a model that always predicts "no event"
would already score ~87%. That is why, from the next step on, we score models with
**ROC-AUC** (and keep accuracy only as a secondary reference).
