---
title: "Step 04 — The zoo, tuned"
subtitle: "Four models compete; cross-validation picks the winner"
params:
  solved: true
format:
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
---

## Where we are

Cumulative snapshot: data + split + recipe are given. The new beat is the heart of
the workshop — a **`workflow_set`** of four candidate models (penalized logistic,
kNN, RBF-SVM, random forest), each **tuned by 5-fold cross-validation**. We rank
them by ROC-AUC (watching PR-AUC too), finalize the winner, and validate it once on
the held-out test set with `last_fit()`.

<!--
Nota docente (IT): step HANDS-ON (~68 min), il CUORE del workshop. I beat
precedenti (load/wrangle/split/recipe) sono "given". Il beat nuovo: definiamo 4
spec con iperparametri `tune()`, li impacchettiamo in un `workflow_set`, e
`workflow_map("tune_grid", ...)` li tuna tutti su 5-fold CV (grid = 8) con DUE
metriche (roc_auc, pr_auc). Il "Your turn" CULMINANTE è il post-tuning:
rank_results -> extract -> select_best -> finalize_workflow -> last_fit
(R/yourturn-04.R con i `___`). Convenzione EVENT-FIRST: `died` è il primo livello,
quindi roc_curve/pr_curve puntano a `.pred_died` senza `event_level`.
PUNTO DIDATTICO (rumore del piccolo N): il glm "vanilla" dello step 03 segnava
~0.88 su UN singolo test split, ma in CV la logistica regolarizzata sta molto più
in basso e il RANDOM FOREST è in testa -> il singolo numero era ottimistico, la CV
è il confronto onesto. Il parallelo `future` è un TOGGLE opzionale (a N=299 il
tuning è di ~secondi; commenta la riga `plan()` e gira lo stesso, in sequenziale).
-->

# Given — data, split, recipe {#sec-given}

From Steps 00–03.

```{r given}
#| code-fold: true
library(here)
library(rio)
library(tidyverse)
library(janitor)
library(tidymodels)
set.seed(123)

hf_raw <- import(here("data-raw", "heart_failure.csv"), setclass = "tibble")

hf <- hf_raw |>
  clean_names() |>
  select(-time) |>
  mutate(
    outcome = factor(
      if_else(death_event == 1, "died", "survived"),
      levels = c("died", "survived"),
    ),
  ) |>
  select(-death_event) |>
  mutate(across(c(anaemia, diabetes, high_blood_pressure, sex, smoking), factor))

# Split ----
data_split <- initial_split(hf, prop = 0.75, strata = outcome)
train <- training(data_split)
test  <- testing(data_split)

# Recipe (no imputation: heart_failure has no missing values) ----
base_rec <- recipe(outcome ~ ., data = train) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())
```

# The zoo — four tunable models {#sec-zoo}

Each spec leaves its key hyper-parameters as `tune()`. We bundle them with the one
recipe into a `workflow_set`, then `workflow_map()` tunes every workflow over a
size-8 grid on 5-fold cross-validation, scoring **both** ROC-AUC and PR-AUC.

```{r tune}
#| code-fold: false
library(workflowsets)
library(future)

# Optional speed-up: parallelize tuning across cores, leaving one for the OS.
# Comment this line out and everything still runs — just sequentially.
plan(multisession, workers = max(1, parallel::detectCores() - 1))
set.seed(123)

# Model specs (hyper-parameters left to tune) ----
penlog_spec <- logistic_reg(penalty = tune(), mixture = tune()) |> set_engine("glmnet")
knn_spec    <- nearest_neighbor(neighbors = tune()) |> set_engine("kknn")   |> set_mode("classification")
svm_spec    <- svm_rbf(cost = tune(), rbf_sigma = tune()) |> set_engine("kernlab") |> set_mode("classification")
rf_spec     <- rand_forest(mtry = tune(), min_n = tune()) |> set_engine("ranger")  |> set_mode("classification")

# Resamples ----
folds <- vfold_cv(train, v = 5, strata = outcome)

# Two metrics, watched together ----
hf_metrics <- metric_set(roc_auc, pr_auc)

# Workflow set ----
wf_set <- workflow_set(
  preproc = list(rec = base_rec),
  models  = list(penlog = penlog_spec, knn = knn_spec, svm = svm_spec, rf = rf_spec),
)

# Tune every workflow ----
wf_res <- wf_set |>
  workflow_map(
    "tune_grid",
    resamples = folds,
    grid      = 8,
    metrics   = hf_metrics,
    verbose   = TRUE,
    seed      = 123,
  )
```

## Compare the candidates

```{r compare}
#| code-fold: false
rank_results(wf_res, rank_metric = "roc_auc", select_best = TRUE)
autoplot(wf_res)
```

Read the ranking honestly. Our Step-03 glm scored high on a *single* test split,
but on cross-validation the **random forest** tops the table while the linear
models trail it — the single test number was optimistic, and the cross-validated
comparison is the one we trust. Both metrics tell a consistent story here.

# Your turn — finalize the winner & validate {#sec-yourturn}

Rank the models, pull the winner's id and best hyper-parameters, **finalize** that
workflow, and run **`last_fit()`** on the original split to score it once on the
held-out test set — reporting ROC-AUC, PR-AUC, and accuracy.

```{r finalize-solved, eval=params$solved, echo=params$solved}
#| code-fold: false
# Finalize & validate ----
best_id     <- rank_results(wf_res, rank_metric = "roc_auc", select_best = TRUE) |>
  dplyr::slice(1) |> dplyr::pull(wflow_id)
best_res    <- extract_workflow_set_result(wf_res, best_id)
best_params <- select_best(best_res, metric = "roc_auc")
final_wf    <- wf_set |> extract_workflow(best_id) |> finalize_workflow(best_params)

final_fit <- last_fit(final_wf, split = data_split, metrics = metric_set(roc_auc, pr_auc, accuracy))
collect_metrics(final_fit)
```

```{r finalize-blank, eval=!params$solved, results="asis", echo=FALSE}
# BLANK mode: echo the `___` skeleton (non-executed) from R/yourturn-04.R.
cat("```r", readLines(here("R", "yourturn-04.R")), "```", sep = "\n")
```

## Test-set ROC & PR curves, then save the winner

```{r curves-and-save, eval=params$solved, echo=params$solved}
#| code-fold: false
preds <- collect_predictions(final_fit)

# Both curves — ROC (ranking) and PR (positive-class focus) ----
preds |> roc_curve(outcome, .pred_died) |> autoplot()
preds |> pr_curve(outcome, .pred_died) |> autoplot()

# Persist the validated workflow + the CV ranking for the report (Step 05) ----
saveRDS(final_fit, here("output", "final_fit.rds"))
saveRDS(
  rank_results(wf_res, rank_metric = "roc_auc", select_best = TRUE),
  here("output", "wf_rank.rds"),
)
```

The cross-validated ranking points at one model; `last_fit()` then reports its
**honest test performance** on both metrics. That single, validated `final_fit`
object is what the next step turns into a self-contained, reproducible report —
without ever re-tuning.
