---
title: "Step 04 — The zoo, tuned"
subtitle: "Four models compete; cross-validation picks the winner"
params:
  solved: true
format:
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
---

## Where we are

Cumulative snapshot: data + split + recipe are given. The new beat is the heart of
the workshop — a **`workflow_set`** of four candidate models (penalized logistic,
kNN, RBF-SVM, random forest), each **tuned by 5-fold cross-validation**. We rank
them by ROC-AUC, finalize the winner, and validate it once on the held-out test set
with `last_fit()`.

<!--
Nota docente (IT): step HANDS-ON (~68 min), il CUORE del workshop. I beat
precedenti (load/wrangle/split/recipe) sono "given". Il beat nuovo: definiamo 4
spec con iperparametri `tune()`, li impacchettiamo in un `workflow_set`, e
`workflow_map("tune_grid", ...)` li tuna tutti su 5-fold CV (grid = 8).
Il "Your turn" CULMINANTE è il post-tuning: rank_results -> extract -> select_best
-> finalize_workflow -> last_fit (R/yourturn-04.R con i `___`).
Wall-clock del tuning misurato: ~18s (dataset piccolo, `plan(multisession)`),
quindi NESSUNA riduzione della griglia svm necessaria.
L'evento è `1_yes` (secondo livello) -> `event_level = "second"` nel roc_curve.
-->

# Given — data, split, recipe {#sec-given}

From Steps 00–03.

```{r given}
#| code-fold: true
library(here)
library(rio)
library(tidyverse)
library(janitor)
library(tidymodels)
set.seed(123)

indo_raw <- import(here("data-raw", "indo_rct.csv"), setclass = "tibble")

indo <- indo_raw |>
  clean_names() |>
  select(outcome, age, risk, gender, sod, rx, type) |>
  filter(!is.na(age)) |>
  mutate(outcome = factor(outcome, levels = c("0_no", "1_yes")))

# Split ----
data_split <- initial_split(indo, prop = 0.75, strata = outcome)
train <- training(data_split)
test  <- testing(data_split)

# Recipe ----
base_rec <- recipe(outcome ~ ., data = train) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())
```

# The zoo — four tunable models {#sec-zoo}

Each spec leaves its key hyper-parameters as `tune()`. We bundle them with the one
recipe into a `workflow_set`, then `workflow_map()` tunes every workflow over a
size-8 grid on 5-fold cross-validation.

```{r tune}
#| code-fold: false
library(workflowsets)
library(future)
plan(multisession)
set.seed(123)

# Model specs (hyper-parameters left to tune) ----
penlog_spec <- logistic_reg(penalty = tune(), mixture = tune()) |> set_engine("glmnet")
knn_spec    <- nearest_neighbor(neighbors = tune()) |> set_engine("kknn")   |> set_mode("classification")
svm_spec    <- svm_rbf(cost = tune(), rbf_sigma = tune()) |> set_engine("kernlab") |> set_mode("classification")
rf_spec     <- rand_forest(mtry = tune(), min_n = tune()) |> set_engine("ranger")  |> set_mode("classification")

# Resamples ----
folds <- vfold_cv(train, v = 5, strata = outcome)

# Workflow set ----
wf_set <- workflow_set(
  preproc = list(rec = base_rec),
  models  = list(penlog = penlog_spec, knn = knn_spec, svm = svm_spec, rf = rf_spec),
)

# Tune every workflow ----
wf_res <- wf_set |>
  workflow_map(
    "tune_grid",
    resamples = folds,
    grid      = 8,
    metrics   = metric_set(roc_auc),
    verbose   = TRUE,
    seed      = 123,
  )
```

## Compare the candidates

```{r compare}
#| code-fold: false
rank_results(wf_res, rank_metric = "roc_auc", select_best = TRUE)
autoplot(wf_res)
```

# Your turn — finalize the winner & validate {#sec-yourturn}

Rank the models, pull the winner's id and best hyper-parameters, **finalize** that
workflow, and run **`last_fit()`** on the original split to score it once on the
held-out test set. The event is `1_yes` (the *second* factor level).

```{r finalize-solved, eval=params$solved}
#| code-fold: false
# Finalize & validate ----
best_id     <- rank_results(wf_res, rank_metric = "roc_auc", select_best = TRUE) |>
  dplyr::slice(1) |> dplyr::pull(wflow_id)
best_res    <- extract_workflow_set_result(wf_res, best_id)
best_params <- select_best(best_res, metric = "roc_auc")
final_wf    <- wf_set |> extract_workflow(best_id) |> finalize_workflow(best_params)

final_fit <- last_fit(final_wf, split = data_split, metrics = metric_set(roc_auc, accuracy))
collect_metrics(final_fit)
```

```{r finalize-blank, eval=!params$solved, results="asis", echo=FALSE}
# BLANK mode: echo the `___` skeleton (non-executed) from R/yourturn-04.R.
cat("```r", readLines(here("R", "yourturn-04.R")), "```", sep = "\n")
```

## Test-set ROC curve & save the winner

```{r roc-and-save, eval=params$solved}
#| code-fold: false
collect_predictions(final_fit) |>
  roc_curve(outcome, .pred_1_yes, event_level = "second") |>
  autoplot()

# Persist the validated workflow + the CV ranking for the report (Step 05) ----
saveRDS(final_fit, here("output", "final_fit.rds"))
saveRDS(
  rank_results(wf_res, rank_metric = "roc_auc", select_best = TRUE),
  here("output", "wf_rank.rds"),
)
```

The cross-validated ranking points at one model; `last_fit()` then reports its
**honest test performance**. That single, validated `final_fit` object is what the
next step turns into a self-contained, reproducible report — without ever
re-tuning.
