---
title: "Step 03 — Logistic spine"
subtitle: "The tidymodels object-map, with a glm baseline"
params:
  solved: true
format:
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
---

## Where we are

Cumulative snapshot: data load + wrangle + EDA are given. The new beat lays down
the **tidymodels object-map** — split the data (`rsample`), build a preprocessing
recipe (`recipes`), specify a model (`parsnip`), glue them in a `workflow`, fit a
plain logistic-regression baseline, and finally score it (`yardstick`).

<!--
Nota docente (IT): step HANDS-ON (~38 min). I beat precedenti (load/wrangle/EDA)
sono "given" e ripiegati. Il beat nuovo (split -> recipe -> spec -> workflow ->
fit) è la "mappa degli oggetti" tidymodels, mostrata risolta. Il "Your turn" di
QUESTO step è SOLO il predict/score finale (`augment` + le due metriche). Grazie
alla convenzione EVENT-FIRST (livello `died` primo), `roc_auc`/`pr_auc` puntano a
`.pred_died` SENZA bisogno di `event_level`. Nota recipe: heart_failure è SENZA
valori mancanti -> niente step di imputazione, la recipe resta a dummy/zv/normalize.
-->

# Given — load, wrangle, EDA {#sec-given}

From Steps 00–02 (table/figure suppressed here to keep the recap short).

```{r given}
#| code-fold: true
library(here)
library(rio)
library(tidyverse)
library(janitor)

hf_raw <- import(here("data-raw", "heart_failure.csv"), setclass = "tibble")

hf <- hf_raw |>
  clean_names() |>
  select(-time) |>
  mutate(
    outcome = factor(
      if_else(death_event == 1, "died", "survived"),
      levels = c("died", "survived"),
    ),
  ) |>
  select(-death_event) |>
  mutate(across(c(anaemia, diabetes, high_blood_pressure, sex, smoking), factor))

glimpse(hf)
```

# The tidymodels object-map {#sec-objectmap}

Split → recipe → spec → workflow → fit. Five objects, one chain. (Heart-failure
has no missing values, so the recipe needs no imputation — just dummy-coding,
zero-variance removal, and normalization.)

```{r objectmap}
#| code-fold: false
library(tidymodels)
set.seed(123)

# Split ----
data_split <- initial_split(hf, prop = 0.75, strata = outcome)
train <- training(data_split)
test  <- testing(data_split)

# Recipe ----
base_rec <- recipe(outcome ~ ., data = train) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

# Spec ----
log_spec <- logistic_reg() |> set_engine("glm")

# Workflow ----
log_wf <- workflow() |>
  add_recipe(base_rec) |>
  add_model(log_spec)

# Fit ----
log_fit <- fit(log_wf, data = train)
log_fit
```

# Your turn — score the baseline on the test set {#sec-yourturn}

Attach predictions to the held-out `test` set with `augment()`, then score **both**
metrics on the event class. Because `died` is the *first* factor level, point
yardstick at the `.pred_died` column — no `event_level` argument needed.

```{r score-solved, eval=params$solved}
#| code-fold: false
# Predict & score (both metrics) ----
hf_metrics <- metric_set(roc_auc, pr_auc)

log_fit |>
  augment(new_data = test) |>
  hf_metrics(truth = outcome, .pred_died)
```

```{r score-blank, eval=!params$solved, results="asis", echo=FALSE}
# BLANK mode: echo the `___` skeleton (non-executed) from R/yourturn-03.R.
cat("```r", readLines(here("R", "yourturn-03.R")), "```", sep = "\n")
```

These two numbers are our **baseline** — but they come from a *single* split of
just ~75 test patients, so do not over-read them yet. In the next step a small zoo
of models competes under **cross-validation**, the honest comparison that tells us
whether anything truly beats this logistic spine.
