---
title: "Step 03 — Logistic spine"
subtitle: "The tidymodels object-map, with a glm baseline"
params:
  solved: true
format:
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
---

## Where we are

Cumulative snapshot: data load + wrangle + EDA are given. The new beat lays down
the **tidymodels object-map** — split the data (`rsample`), build a preprocessing
recipe (`recipes`), specify a model (`parsnip`), glue them in a `workflow`, fit a
plain logistic-regression baseline, and finally score it (`yardstick`).

<!--
Nota docente (IT): step HANDS-ON (~38 min). I beat precedenti (load/wrangle/EDA)
sono "given" e ripiegati. Il beat nuovo (split -> recipe -> spec -> workflow ->
fit) è la "mappa degli oggetti" tidymodels, mostrata risolta. Il "Your turn" di
QUESTO step è SOLO il predict/score finale (`augment` + `roc_auc`), in
R/yourturn-03.R con i `___`. L'evento è `1_yes` (secondo livello) -> ovunque si
punteggi serve `event_level = "second"`.
Nota dati: `sod` e `type` sono quasi collineari (sod==0_no <=> type==0_no SOD);
glm può emettere un warning di rank-deficiency, soppresso da execute.warning:false.
Non rompe nulla e non richiede di togliere colonne.
-->

# Given — load, wrangle, EDA {#sec-given}

From Steps 00–02 (table/figure suppressed here to keep the recap short).

```{r given}
#| code-fold: true
library(here)
library(rio)
library(tidyverse)
library(janitor)

indo_raw <- import(here("data-raw", "indo_rct.csv"), setclass = "tibble")

indo <- indo_raw |>
  clean_names() |>
  select(outcome, age, risk, gender, sod, rx, type) |>
  filter(!is.na(age)) |>
  mutate(outcome = factor(outcome, levels = c("0_no", "1_yes")))

glimpse(indo)
```

# The tidymodels object-map {#sec-objectmap}

Split → recipe → spec → workflow → fit. Five objects, one chain.

```{r objectmap}
#| code-fold: false
library(tidymodels)
set.seed(123)

# Split ----
data_split <- initial_split(indo, prop = 0.75, strata = outcome)
train <- training(data_split)
test  <- testing(data_split)

# Recipe ----
base_rec <- recipe(outcome ~ ., data = train) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

# Spec ----
log_spec <- logistic_reg() |> set_engine("glm")

# Workflow ----
log_wf <- workflow() |>
  add_recipe(base_rec) |>
  add_model(log_spec)

# Fit ----
log_fit <- fit(log_wf, data = train)
log_fit
```

# Your turn — score the baseline on the test set {#sec-yourturn}

Attach predictions to the held-out `test` set with `augment()`, then compute the
**ROC-AUC** on the event class. The event is `1_yes` (the *second* factor level),
so score `.pred_1_yes` with `event_level = "second"`.

```{r score-solved, eval=params$solved}
#| code-fold: false
# Predict & score ----
log_fit |>
  augment(new_data = test) |>
  roc_auc(truth = outcome, .pred_1_yes, event_level = "second")
```

```{r score-blank, eval=!params$solved, results="asis", echo=FALSE}
# BLANK mode: echo the `___` skeleton (non-executed) from R/yourturn-03.R.
cat("```r", readLines(here("R", "yourturn-03.R")), "```", sep = "\n")
```

The held-out ROC-AUC is our **baseline**. In the next step we let a small zoo of
models compete — tuned by cross-validation — and see whether any beats it.
