---
title: "Practical AI for Medical Data — a reproducible model report"
subtitle: "Predicting post-ERCP pancreatitis in the indomethacin RCT"
date: today
params:
  solved: true
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

<!--
Nota docente (IT): step DEMO (~13 min). Questo è il DELIVERABLE riproducibile.
NON ri-tuna nulla: ricarica `output/final_fit.rds` e `output/wf_rank.rds`
(prodotti allo step 04) e ne presenta i risultati. Tutto data-driven dai due rds,
così il report resta onesto ("no pre-baked shown as live") e veloce (<90s).
-->

## What this report does

This is the reproducible deliverable of the workshop. It does **not** re-train or
re-tune anything: it reloads the model we validated in Step 04 (`final_fit.rds`)
and the cross-validation ranking (`wf_rank.rds`), and reports them. Re-rendering
this file on any machine with the pinned environment reproduces every number and
figure below.

```{r load}
library(here)
library(rio)
library(tidyverse)
library(janitor)
library(tidymodels)
library(gtsummary)

# Reload the validated artifacts from Step 04 (no tuning happens here) ----
final_fit <- readRDS(here("output", "final_fit.rds"))
wf_rank   <- readRDS(here("output", "wf_rank.rds"))

# Reload the analysis cohort for the descriptive table ----
indo <- import(here("data-raw", "indo_rct.csv"), setclass = "tibble") |>
  clean_names() |>
  select(outcome, age, risk, gender, sod, rx, type) |>
  filter(!is.na(age)) |>
  mutate(outcome = factor(outcome, levels = c("0_no", "1_yes")))
```

## The cohort

The analysis cohort is the indomethacin RCT for post-ERCP pancreatitis
(`medicaldata::indo_rct`): `r nrow(indo)` patients, outcome `1_yes` (pancreatitis)
in `r scales::percent(mean(indo$outcome == "1_yes"), accuracy = 0.1)` of them. The
table below summarises the six candidate predictors, stratified by outcome.

```{r baseline}
indo |>
  tbl_summary(
    by = outcome,
    include = c(age, risk, gender, sod, rx, type),
  )
```

## The winning model

Four model families (penalized logistic regression, k-nearest neighbours,
RBF-SVM, random forest) competed, each tuned over a grid on 5-fold
cross-validation. The ranking by cross-validated ROC-AUC:

```{r ranking}
model_label <- c(
  rec_penlog = "Penalized logistic (glmnet)",
  rec_knn    = "k-nearest neighbours",
  rec_svm    = "RBF support vector machine",
  rec_rf     = "Random forest"
)

wf_rank |>
  transmute(
    Model      = model_label[wflow_id],
    `CV ROC-AUC` = round(mean, 3),
    `Std. error` = round(std_err, 3),
    Rank       = rank,
  )
```

```{r winner-numbers}
winner_id  <- wf_rank |> dplyr::slice_min(rank, n = 1) |> dplyr::pull(wflow_id)
cv_auc     <- wf_rank |> dplyr::slice_min(rank, n = 1) |> dplyr::pull(mean)
test_auc   <- collect_metrics(final_fit) |>
  dplyr::filter(.metric == "roc_auc") |> dplyr::pull(.estimate)
test_acc   <- collect_metrics(final_fit) |>
  dplyr::filter(.metric == "accuracy") |> dplyr::pull(.estimate)
```

The winner is **`r model_label[winner_id]`**. Its cross-validated ROC-AUC was
**`r round(cv_auc, 3)`**; refit on the training split and scored once on the
held-out test set, it reached a test ROC-AUC of **`r round(test_auc, 3)`** and an
accuracy of **`r round(test_acc, 3)`**. The drop from CV to test is the honest
cost of a small, imbalanced cohort — and exactly why we report a single,
pre-committed test evaluation rather than the optimistic CV number alone.

## Test-set ROC curve

```{r roc-curve}
collect_predictions(final_fit) |>
  roc_curve(outcome, .pred_1_yes, event_level = "second") |>
  autoplot()
```

## Reproducibility

Every result above is reproducible by construction:

- **Pinned environment.** Package versions are locked with `renv`; running
  `renv::snapshot()` records them in `renv.lock`, and `renv::restore()` rebuilds
  the exact library on another machine.
- **Stable paths.** All file access goes through `here::here()`, anchored to this
  step's `.here` sentinel, so the report renders identically regardless of the
  working directory.
- **No live re-tuning.** The model was tuned and validated once (Step 04) and
  persisted to `output/final_fit.rds`; this report only reloads and presents it,
  so re-rendering is fast and deterministic.
- **One outcome convention.** The event is `1_yes` (the second factor level), and
  every metric is scored with `event_level = "second"` — consistently, end to end.

That is the whole point of the workshop: a clinical model you can hand to someone
else and have them reproduce, number for number.
