---
title: "Practical AI for Medical Data — a reproducible model report"
subtitle: "Predicting in-hospital death after a heart-failure episode"
date: today
params:
  solved: true
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

<!--
Nota docente (IT): step DEMO (~13 min). Questo è il DELIVERABLE riproducibile.
NON ri-tuna nulla: ricarica `output/final_fit.rds` e `output/wf_rank.rds`
(prodotti allo step 04) e ne presenta i risultati. Tutto data-driven dai due rds,
così il report resta onesto ("no pre-baked shown as live") e veloce (<90s).
-->

## What this report does

This is the reproducible deliverable of the workshop. It does **not** re-train or
re-tune anything: it reloads the model we validated in Step 04 (`final_fit.rds`)
and the cross-validation ranking (`wf_rank.rds`), and reports them. Re-rendering
this file on any machine with the pinned environment reproduces every number and
figure below.

```{r load}
library(here)
library(rio)
library(tidyverse)
library(janitor)
library(tidymodels)
library(gtsummary)

# Reload the validated artifacts from Step 04 (no tuning happens here) ----
final_fit <- readRDS(here("output", "final_fit.rds"))
wf_rank   <- readRDS(here("output", "wf_rank.rds"))

# Reload the analysis cohort for the descriptive table ----
hf <- import(here("data-raw", "heart_failure.csv"), setclass = "tibble") |>
  clean_names() |>
  select(-time) |>
  mutate(
    outcome = factor(
      if_else(death_event == 1, "died", "survived"),
      levels = c("died", "survived"),
    ),
  ) |>
  select(-death_event) |>
  mutate(across(c(anaemia, diabetes, high_blood_pressure, sex, smoking), factor))
```

## The cohort

The analysis cohort is the heart-failure clinical records (Chicco & Jurman 2020):
`r nrow(hf)` patients, of whom
`r scales::percent(mean(hf$outcome == "died"), accuracy = 0.1)` died during
follow-up. The table below summarises the eleven candidate predictors, stratified
by outcome.

```{r baseline}
hf |>
  tbl_summary(
    by = outcome,
    include = c(
      age, sex, ejection_fraction, serum_creatinine, serum_sodium,
      creatinine_phosphokinase, platelets, anaemia, diabetes,
      high_blood_pressure, smoking,
    ),
  )
```

## The winning model

Four model families (penalized logistic regression, k-nearest neighbours,
RBF-SVM, random forest) competed, each tuned over a grid on 5-fold
cross-validation. The ranking by cross-validated AUC-ROC:

```{r ranking}
model_label <- c(
  rec_penlog = "Penalized logistic (glmnet)",
  rec_knn    = "k-nearest neighbours",
  rec_svm    = "RBF support vector machine",
  rec_rf     = "Random forest"
)

wf_rank |>
  dplyr::filter(.metric == "roc_auc") |>
  transmute(
    Model        = model_label[wflow_id],
    `CV AUC-ROC` = round(mean, 3),
    `Std. error` = round(std_err, 3),
    Rank         = rank,
  )
```

```{r winner-numbers}
winner_id <- wf_rank |>
  dplyr::filter(.metric == "roc_auc") |>
  dplyr::slice_min(rank, n = 1) |> dplyr::pull(wflow_id)
cv_auc <- wf_rank |>
  dplyr::filter(.metric == "roc_auc") |>
  dplyr::slice_min(rank, n = 1) |> dplyr::pull(mean)

test_metrics <- collect_metrics(final_fit)
get_metric <- function(m) {
  test_metrics |> dplyr::filter(.metric == m) |> dplyr::pull(.estimate)
}
test_roc <- get_metric("roc_auc")
test_pr  <- get_metric("pr_auc")
test_acc <- get_metric("accuracy")
```

The winner is **`r model_label[winner_id]`**. Its cross-validated AUC-ROC was
**`r round(cv_auc, 3)`**; refit on the training split and scored once on the
held-out test set, it reached a test AUC-ROC of **`r round(test_roc, 3)`**, a
AUC-PR of **`r round(test_pr, 3)`**, and an accuracy of
**`r round(test_acc, 3)`**. The gap between the cross-validated and the
single-split test numbers is the honest noise of a small cohort (~75 test
patients) — which is exactly why we pre-commit to one held-out evaluation and read
it *alongside* the cross-validation, never a single number on its own.

## Test-set ROC and PR curves

```{r curves}
preds <- collect_predictions(final_fit)

preds |> roc_curve(outcome, .pred_died) |> autoplot()
preds |> pr_curve(outcome, .pred_died) |> autoplot()
```

## Reproducibility

Every result above is reproducible by construction:

- **Pinned environment.** Package versions are locked with `renv`; running
  `renv::snapshot()` records them in `renv.lock`, and `renv::restore()` rebuilds
  the exact library on another machine.
- **Stable paths.** All file access goes through `here::here()`, anchored to this
  step's `.here` sentinel, so the report renders identically regardless of the
  working directory.
- **No live re-tuning.** The model was tuned and validated once (Step 04) and
  persisted to `output/final_fit.rds`; this report only reloads and presents it,
  so re-rendering is fast and deterministic.
- **One outcome convention.** The event is `died` (the *first* factor level), so
  every metric scores `.pred_died` as the positive class by default — consistently,
  end to end.

That is the whole point of the workshop: a clinical model you can hand to someone
else and have them reproduce, number for number.
