---
title: "Practical AI for Medical Data — a reproducible model report"
subtitle: "Predicting in-hospital death after a heart-failure episode"
date: today
format:
  html:
    embed-resources: true
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

## What this report does

This is the reproducible deliverable of the workshop. It **runs the entire analysis as it
renders** — split → preprocess → tune four model families on cross-validation → finalize the
winner → validate once on the held-out test set — and then reports the result. Nothing is
typed in by hand and nothing is reloaded from a private cache: **every number and figure below
was produced by the same render that prints it.** Re-render on any machine with the pinned
environment and you reproduce the *analysis*, not just the document.

```{r setup}
library(here)
library(rio)
library(tidyverse)
library(janitor)
library(tidymodels)
library(gtsummary)
```

## The cohort

```{r cohort}
# Load & wrangle exactly as in Steps 00-01 ----
hf <- import(here("data-raw", "heart_failure.csv"), setclass = "tibble") |>
{{frag:wrangle-tail}}
```

The analysis cohort is the heart-failure clinical records (Chicco & Jurman 2020):
`r nrow(hf)` patients, of whom
`r scales::percent(mean(hf$outcome == "died"), accuracy = 0.1)` died during follow-up. The
table below summarises the eleven candidate predictors, stratified by outcome.

```{r baseline}
hf |>
  tbl_summary(
    by = outcome,
    include = c(
      age, sex, ejection_fraction, serum_creatinine, serum_sodium,
      creatinine_phosphokinase, platelets, anaemia, diabetes,
      high_blood_pressure, smoking,
    ),
  )
```

## Split and preprocess

We hold out a stratified 25% test set that **no tuning ever sees**, and build the same recipe
used throughout the workshop (dummy-code, drop zero-variance, normalize — no imputation, the
data has no missing values).

```{r split-recipe}
# Split: a held-out test set the tuning never touches ----
set.seed(123)
{{frag:split}}

# Recipe ----
{{frag:recipe}}
```

## Tune four model families on cross-validation

Each spec leaves its hyperparameters as `tune()`. We bundle the one recipe with the four
specs into a `workflow_set`, then tune every workflow over a shared 5-fold cross-validation,
scoring **both** AUC-ROC and AUC-PR.

```{r tune}
# Model specs (hyperparameters left to tune) ----
{{frag:specs}}

# One shared resampling scheme and metric set ----
folds      <- vfold_cv(train, v = 5, strata = outcome)
hf_metrics <- metric_set(roc_auc, pr_auc)

# Bundle {one recipe} x {four specs}, then tune every workflow ----
wf_set <- workflow_set(
  preproc = list(rec = base_rec),
  models  = list(penlog = penlog_spec, knn = knn_spec, svm = svm_spec, rf = rf_spec),
)

wf_res <- wf_set |>
  workflow_map(
    "tune_grid",
    resamples = folds,
    grid      = 8,
    metrics   = hf_metrics,
    seed      = 123,
    verbose   = FALSE,
  )
```

The ranking by cross-validated AUC-ROC (the honest comparison — it never touches the test
set):

```{r ranking}
wf_rank <- rank_results(wf_res, rank_metric = "roc_auc", select_best = TRUE)

model_label <- c(
  rec_penlog = "Penalized logistic (glmnet)",
  rec_knn    = "k-nearest neighbours",
  rec_svm    = "RBF support vector machine",
  rec_rf     = "Random forest"
)

wf_rank |>
  dplyr::filter(.metric == "roc_auc") |>
  transmute(
    Model        = model_label[wflow_id],
    `CV AUC-ROC` = round(mean, 3),
    `Std. error` = round(std_err, 3),
    Rank         = rank,
  )
```

## Finalize the winner and validate once

We take the top-ranked workflow, lock in its best hyperparameters, refit on the full training
set, and score it **once** on the untouched test set.

```{r finalize}
best_id     <- wf_rank |>
  dplyr::filter(.metric == "roc_auc") |>
  dplyr::slice_min(rank, n = 1) |> dplyr::pull(wflow_id)
best_res    <- extract_workflow_set_result(wf_res, best_id)
best_params <- select_best(best_res, metric = "roc_auc")
final_wf    <- wf_set |> extract_workflow(best_id) |> finalize_workflow(best_params)

final_fit <- last_fit(
  final_wf,
  split   = data_split,
  metrics = metric_set(roc_auc, pr_auc, accuracy),
)

cv_auc       <- wf_rank |>
  dplyr::filter(.metric == "roc_auc") |>
  dplyr::slice_min(rank, n = 1) |> dplyr::pull(mean)
test_metrics <- collect_metrics(final_fit)
get_metric   <- function(m) {
  test_metrics |> dplyr::filter(.metric == m) |> dplyr::pull(.estimate)
}
test_roc <- get_metric("roc_auc")
test_pr  <- get_metric("pr_auc")
test_acc <- get_metric("accuracy")
```

The winner is **`r model_label[best_id]`**. Its cross-validated AUC-ROC was
**`r round(cv_auc, 3)`**; refit on the training split and scored once on the held-out test
set, it reached a test AUC-ROC of **`r round(test_roc, 3)`**, an AUC-PR of
**`r round(test_pr, 3)`**, and an accuracy of **`r round(test_acc, 3)`**. The gap between the
cross-validated and the single-split test numbers is the honest noise of a small cohort
(~`r nrow(test)` test patients) — which is exactly why we pre-commit to one held-out
evaluation and read it *alongside* the cross-validation, never a single number on its own.

## Test-set ROC and PR curves

```{r curves}
preds <- collect_predictions(final_fit)

preds |> roc_curve(outcome, .pred_died) |> autoplot()
preds |> pr_curve(outcome, .pred_died) |> autoplot()
```

## Reproducibility

Every result above is reproducible by construction:

- **One render, one analysis.** This report *executes* the split → tune → finalize →
  `last_fit` pipeline as it renders, so the printed numbers and the code that produced them
  are guaranteed to be the same object — they cannot drift apart, because there is no
  hand-made cache to fall out of date.
- **Fixed randomness.** `set.seed(123)` makes the split, the folds, and the tuning grid
  deterministic, so the same machine re-renders to the same numbers.
- **Pinned environment.** Package versions are locked with `renv`; `renv::snapshot()` records
  them in `renv.lock`, and `renv::restore()` rebuilds the exact library on another machine.
- **Stable paths.** All file access goes through `here::here()`, anchored to this step's
  `.here` sentinel, so the report renders identically regardless of the working directory.
- **One outcome convention.** The event is `died` (the *first* factor level), so every metric
  scores `.pred_died` as the positive class by default — consistently, end to end.

That is the whole point: a clinical model someone else can re-run — and reproduce the
*analysis*, number for number, not just the document. When an analysis grows too heavy to
re-run on every render, the next step is not a hand-made `.rds` cache but a **pipeline**
(`targets`, the Advanced capstone): it caches *with* inspectable, hash-checked intermediate
targets — each one re-derivable and auditable — and compiles this very report itself.
