---
title: "Step 01 — Open the black box"
subtitle: "Permutation VIMP + agnostic SHAP, anchored on a logistic regression"
params:
  solved: true
format:
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
---

## Where we are

We have a validated random forest from the Basic workshop, but a probability is not an
explanation. This step opens the box with two **model-agnostic** tools. First,
**permutation variable importance (VIMP)**: shuffle one predictor, watch how far ROC-AUC
falls, and read that drop as how much the model *relies* on that feature. Second,
**SHAP**, which attributes a single prediction across its features. To trust SHAP before
we point it at the forest, we anchor it on a **logistic regression** — an interpretable
model whose coefficient signs SHAP must recover. The payoff is agnosticism: the same call
explains the linear anchor and the forest alike; only the model object changes.

<!--
Nota docente (IT): step HANDS-ON (~48 min). Cinque messaggi chiave:
1. VIMP = permuta-una-feature -> il calo di performance = quanto il modello DIPENDE da
   quella feature. NON e' causalita', NON e' il coefficiente, NON e' un p-value
   (formativa min-24). E' "reliance" e basta.
2. L'ANCORA logistica e' un dispositivo di sanity: SHAP, applicato a un modello lineare,
   DEVE far riemergere i segni/le grandezze dei coefficienti (formativa min-38). Se non
   lo facesse, non ci fideremmo di SHAP sul random forest.
3. Le MANOPOLE di SHAP: la dimensione del background (`bg_X`), QUALI righe spieghi (`X`),
   ed esatto-vs-campionato (formativa min-46). Con 11 feature, permshap() passa in
   automatico alla versione campionata: lo segnaliamo, resta il check piu' pulito su un
   modello lineare.
4. Il payoff dell'agnosticismo: STESSA chiamata, cambi solo il modello. Il "Your turn"
   prende esattamente la chiamata dell'ancora e la punta sul random forest.
5. CAVEAT tecnico (vip + workflow): vip "spacchetta" il workflow e passa al pred_wrapper
   l'oggetto-motore nudo (ranger), su cui predict() non parla tidymodels. Quindi il
   wrapper instrada SEMPRE attraverso il workflow catturato (`rf_wf`), ignorando `object`.
   E' un trucco idiomatico, non un bug: lo commentiamo nel chunk.
Sanity attesa: ejection_fraction / serum_creatinine in alto nella VIMP (soglie/effetti
non-monotoni noti del dataset). Confermato in build: ejection_fraction > age >
serum_creatinine.
-->

## Given — cohort, split, anchor inputs {#sec-given}

Rebuilt exactly as in Basic: `clean_names()` first, drop `time`, event-first `outcome`,
binary predictors as factors. We also reload the Basic random forest as-is.

```{r given}
#| code-fold: true
library(here)
library(rio)
library(tidyverse)
library(janitor)
library(tidymodels)
library(vip)
library(kernelshap)
library(shapviz)
set.seed(123)

# Rebuild the cohort exactly as in Basic (clean_names first) ----
hf_tbl <- import(here("data-raw", "heart_failure.csv"), setclass = "tibble") |>
  clean_names() |>
  select(-time) |>
  mutate(
    outcome = factor(
      if_else(death_event == 1, "died", "survived"),
      levels = c("died", "survived"),
    ),
  ) |>
  select(-death_event) |>
  mutate(across(c(anaemia, diabetes, high_blood_pressure, sex, smoking), factor))

# A train split for the anchor + a SHAP background ----
data_split <- initial_split(hf_tbl, prop = 0.75, strata = outcome)
train <- training(data_split)

# Reload the validated Basic random forest (no retraining) ----
rf_wf <- extract_workflow(readRDS(here("model", "final_fit.rds")))
```

## Permutation VIMP — what the forest relies on {#sec-vimp}

We fit the **logistic anchor** (it earns its keep in the next section) and then measure
permutation importance on the forest. VIMP shuffles one predictor at a time and records
how much ROC-AUC drops: a big drop means the model leans hard on that feature. Read it as
*reliance*, not as causation, not as a coefficient, and not as a p-value.

```{r anchor-vimp}
#| code-fold: false
# Logistic ANCHOR: an interpretable model whose coefficient signs SHAP must recover ----
anchor_rec <- recipe(outcome ~ ., data = train) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())
log_wf  <- workflow() |> add_recipe(anchor_rec) |> add_model(logistic_reg() |> set_engine("glm"))
log_fit <- fit(log_wf, train)

# Permutation VIMP (agnostic): shuffle a predictor -> how much does ROC-AUC drop? ----
# NOTE: vip unwraps the workflow and hands the bare engine (ranger) to pred_wrapper, so we
# ignore `object` and route the prediction through the captured workflow `rf_wf` instead.
vimp_pred <- function(object, newdata) predict(rf_wf, newdata, type = "prob")$.pred_died
set.seed(1)
vip(
  rf_wf,
  method       = "permute",
  train        = train,
  target       = "outcome",
  metric       = "roc_auc",
  pred_wrapper = vimp_pred,
  nsim         = 10,
  event_level  = "first",
)
```

The top of the plot is dominated by **ejection_fraction** and **serum_creatinine** (with
**age**): exactly the features clinicians watch, and the ones with known threshold /
non-monotone effects on survival. That is reliance the forest learned from data — it says
nothing, on its own, about *why* or in *which direction*.

## SHAP on the anchor — the sanity check {#sec-shap}

SHAP attributes a single prediction across its features. Three knobs govern it: the
**background** `bg_X` (the reference distribution SHAP integrates over), **which rows** you
explain via `X`, and **exact vs. sampling** (with 11 features `permshap()` switches to its
sampling variant automatically). We run it first on the *linear anchor*, where the answer
is known: the SHAP attributions must echo the fitted logistic coefficients.

```{r shap-anchor}
#| code-fold: false
# Agnostic SHAP on the ANCHOR. kernelshap's pred_fun signature is function(object, X, ...).
pred_fun <- function(object, X) predict(object, X, type = "prob")$.pred_died
bg <- train |> select(-outcome)

# permshap() = (near-)EXACT Shapley values -> the cleanest sanity check on a linear model ----
set.seed(2)
ks_log <- permshap(log_fit, X = bg[1, ], bg_X = bg, pred_fun = pred_fun)
shapviz(ks_log) |> sv_waterfall()

# Why it reassures: the SHAP signs must echo the logistic coefficients ----
tidy(log_fit) |> dplyr::arrange(dplyr::desc(abs(estimate)))
```

Compare the two outputs. The features SHAP pushes hardest in the waterfall are the same
ones with the largest coefficient magnitudes — **ejection_fraction**, **age**,
**serum_creatinine** lead both — and each attribution's direction is consistent with that
patient's (standardized) feature value times its fitted coefficient. SHAP recovers the
linear model it was given, so we can trust the same machinery on a model we *cannot* read
off coefficients.

## Your turn — same call, swap the model {#sec-yourturn}

Here is the agnosticism payoff. The forest has no coefficients, but the explainer does not
care: it only needs a `pred_fun` and a background. Take the anchor's exact call and point
it at the random forest — **only the model object changes**. (We use `kernelshap()` for the
forest: at 11 features its hybrid strategy is the practical choice; `permshap()`'s exact
enumeration is what we reserved for the linear anchor.)

```{r yourturn-01-solved, eval=params$solved}
#| code-fold: false
# Your turn (solved): point the SAME explainer at the RANDOM FOREST. Only the model changed.
set.seed(3)
ks_rf <- kernelshap(rf_wf, X = bg[1, ], bg_X = bg, pred_fun = pred_fun)
shapviz(ks_rf) |> sv_waterfall()
```

```{r yourturn-01-blank, eval=!params$solved, results="asis", echo=FALSE}
cat("```r", readLines(here("R", "yourturn-01.R")), "```", sep = "\n")
```

The waterfall for the forest reads the same way as the anchor's — a per-feature decomposition
of one prediction — but now it explains a model whose internal logic is opaque. That is the
whole point: one agnostic interface, any model behind it.
