---
title: "Step 00 — Recap & setup"
subtitle: "Reload the Basic model; warm up torch"
params:
  solved: true
format:
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
---

<!-- params$solved unused here: step 00 is DEMO-only (no blanks). The flag stays for uniformity with hands-on steps. -->

## Where we are

We reopen the **validated model from the Basic workshop** — a random forest predicting
in-hospital death after heart failure — and warm up `torch` so deep learning is instant
later. No retraining happens here: we simply confirm the workflow loads cleanly and
produces well-formed probability predictions.

<!--
Nota docente (IT): Step DEMO (~10 min). Tre messaggi chiave:
1. Ogni step è uno snapshot cumulativo autonomo — `.here` ancora `here::here()` a questa cartella,
   non alla root del workshop. Il rendering va fatto DALLA ROOT del workshop (così renv si attiva).
2. Il wrangle ripete esattamente la pipeline del Basic: `clean_names()` PRIMA, poi drop `time`,
   poi `outcome` factor con "died" come livello-evento, poi predittori binari a factor.
   Non toccare `death_event` prima di `clean_names()` — nel CSV grezzo è `DEATH_EVENT`.
3. `torch_tensor(1)` preriscalda il backend ora; senza questo, il primo tensore in demo causa
   un'attesa di 3-5 s inaspettata.
-->

## Reload the Basic workflow

The `.rds` contains a finalised `tidymodels` workflow — already fitted on the full Basic
training set. We extract it and run a single-row prediction to confirm nothing is broken.

```{r recap}
#| code-fold: false
library(here)
library(tidymodels)

# Reload the Basic-validated workflow (no retraining) ----
final_fit <- readRDS(here("model", "final_fit.rds"))
fitted_wf <- extract_workflow(final_fit)

# Rebuild the modelling cohort exactly as in Basic (clean_names first!) ----
hf_tbl <- rio::import(here("data-raw", "heart_failure.csv"), setclass = "tibble") |>
  janitor::clean_names() |>
  select(-time) |>
  mutate(
    outcome = factor(if_else(death_event == 1, "died", "survived"),
                     levels = c("died", "survived")),
  ) |>
  select(-death_event) |>
  mutate(across(c(anaemia, diabetes, high_blood_pressure, sex, smoking), factor))

# One-row prediction on the finalised model (sanity-check, no retraining) ----
predict(fitted_wf, hf_tbl[1, ], type = "prob")
```

The two columns `.pred_died` and `.pred_survived` sum to 1; the random forest is alive
and ready.

## Pre-warm torch

Loading `torch` the first time in a session triggers a backend initialisation that can
pause for several seconds. We do it here — outside a demo — so it is invisible later.

```{r warm-torch}
#| code-fold: false
library(torch)
torch_tensor(1)             # forces the backend to load now, not mid-demo
torch::cuda_is_available()  # FALSE here; TRUE on the GPU teaching machine
```

The tensor printed above confirms the LibTorch backend is resident in memory.
`cuda_is_available()` returns `FALSE` on this CPU build machine; it returns `TRUE` in
the classroom on the NVIDIA/CUDA teaching machine.
