---
title: "Capstone report — compiled by the `targets` pipeline"
subtitle: "The whole workshop arc, reproduced live and read straight from the DAG"
format:
  html:
    embed-resources: true
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

```{=html}
<!--MLT-REPORT-START-->
```

<!--
Nota docente (IT): questo e' il REPORT COMPILATO DALLA PIPELINE. NON ri-esegue l'analisi
inline: legge i TARGET INTERMEDI con tar_read(). tar_quarto() in _targets.R rileva ogni
tar_read() e crea le dipendenze, cosi' il report e' un target a valle, riproducibile e
auditabile. Cambia un input a monte -> si invalidano e ricompilano SOLO i target a valle;
altrimenti e' skip. In aula: tar_visnetwork(FALSE) mostra il grafo; tocca un input e fai
vedere che ricomputa solo quel ramo.
-->

This report is **compiled by the `targets` pipeline itself** (`tarchetypes::tar_quarto()`).
It does **not** re-run the analysis inline — it **reads the intermediate targets** with
`tar_read()`, so what you see is, by construction, exactly what the pipeline produced upstream.
The whole workshop arc — explore, compare the base techniques, the forest's importance and SHAP,
a deep-net learning curve, and the typed LLM extraction — is **one `tar_make()`** away. Change one
input and only its downstream targets rebuild; change nothing and the whole thing skips.

```{r setup}
library(targets)
library(ggplot2)
library(dplyr)
library(shapviz)
```

## 1 · Exploration — who is in the cohort

```{r explore}
eda <- tar_read(eda)
eda$balance
```

```{r explore-plot}
ggplot(eda$separation, aes(x = mean_z, y = reorder(variable, mean_z), fill = outcome)) +
  geom_col(position = "dodge") +
  labs(
    title = "How each numeric predictor separates the two outcomes",
    subtitle = "Standardised group means (z) — read from the `eda` target",
    x = "Mean (standardised)", y = NULL, fill = NULL,
  )
```

The classes are imbalanced (more survivors), and `ejection_fraction`, `serum_creatinine`, and `age`
already pull the two groups apart before any model is fit.

## 2 · Base techniques — the four Basic learners, compared

```{r base-compare}
base_compare <- tar_read(base_compare)
base_compare
```

```{r base-plot}
ggplot(base_compare, aes(x = roc_auc, y = reorder(model, roc_auc))) +
  geom_col() +
  geom_errorbarh(aes(xmin = roc_auc - std_err, xmax = roc_auc + std_err), height = 0.2) +
  labs(
    title = "5-fold CV ROC-AUC of the base techniques",
    subtitle = "penalised logistic, kNN, SVM, random forest — read from `base_compare`",
    x = "ROC-AUC (mean ± std. error)", y = NULL,
  )
```

The random forest is the robust winner — the model the rest of the pipeline keeps.

## 3 · Permutation VIMP — what the forest relies on

```{r vimp}
vimp <- tar_read(vimp)
vimp
```

```{r vimp-plot}
vimp |>
  mutate(Variable = stats::reorder(Variable, Importance)) |>
  ggplot(aes(x = Importance, y = Variable)) +
  geom_col() +
  labs(
    title = "Permutation VIMP of the random forest",
    subtitle = "Value of the `vimp` target — compiled by the pipeline",
    x = "Drop in ROC-AUC when permuted", y = NULL,
  )
```

`ejection_fraction`, `serum_creatinine`, and `age` lead — the reliance ordering the forest learned.

## 4 · Agnostic SHAP — the same forest, model-agnostic

```{r shap-plot}
shap <- tar_read(shap)
sv_importance(shap, kind = "beeswarm")
```

Kernel SHAP, treating the forest as a black box, recovers the same drivers — agreement between two
independent explainers, both read from the DAG.

## 5 · A deep-net learning curve

```{r curve-plot}
mlp_curve <- tar_read(mlp_curve)
ggplot(mlp_curve, aes(x = epoch, y = valid_loss)) +
  geom_line() +
  geom_point() +
  labs(
    title = "An MLP learning the cohort: validation loss per epoch",
    subtitle = "Value of the `mlp_curve` target — a live brulee/torch fit",
    x = "Epoch", y = "Validation loss",
  )
```

Validation loss falls to a minimum, then creeps back up as the net starts to overfit; early stopping
halts training a few epochs later — the U and the early-stopping rule from §02, now on the tabular
cohort and re-run by the pipeline.

## 6 · Typed LLM extraction — the ETL, one row per note

```{r records}
records <- tar_read(records)
records
```

Each clinical note is reduced to a validated, typed record — the `ellmer` ETL, re-run by the
pipeline alongside everything else.

---

Because this document is a downstream **target**, re-rendering it is governed by input hashes:
identical inputs ⇒ an all-skip rebuild, the signature of reproducibility. Touch one upstream input
and `tar_visnetwork()` shows exactly which branch — and only that branch — goes stale.

```{=html}
<!--MLT-REPORT-END-->
```
