---
title: "Explanation report — compiled by the `targets` pipeline"
subtitle: "Permutation VIMP of the reloaded Basic random forest"
format:
  html:
    embed-resources: true
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

<!--
Nota docente (IT): questo e' il REPORT COMPILATO DALLA PIPELINE (feedback #13). NON
ri-esegue l'analisi inline: legge i TARGET INTERMEDI con tar_read(). tar_quarto() in
_targets.R rileva tar_read(explanation) e crea la dipendenza, cosi' il report e' un
target a valle, riproducibile e auditabile. Cambia un input a monte -> il report si
invalida e si ricompila; altrimenti e' skip.
-->

This report is **compiled by the `targets` pipeline itself** (`tarchetypes::tar_quarto()`).
It does **not** re-run the analysis inline — it **reads the intermediate targets**
(`tar_read(explanation)`), so what you see here is, by construction, exactly what the
pipeline produced upstream. Change an input and the report is rebuilt; change nothing and
it is skipped.

```{r read-target}
library(targets)
library(ggplot2)

# Read the upstream intermediate target — NOT a re-computation ----
explanation <- tar_read(explanation)
explanation
```

## Permutation VIMP — read straight from the target

```{r plot-target}
explanation |>
  dplyr::mutate(Variable = stats::reorder(Variable, Importance)) |>
  ggplot(aes(x = Importance, y = Variable)) +
  geom_col() +
  labs(
    title = "Permutation VIMP of the reloaded Basic random forest",
    subtitle = "Value of the `explanation` target — compiled by the pipeline",
    x = "Drop in AUC-ROC when permuted",
    y = NULL,
  )
```

`ejection_fraction`, `serum_creatinine`, and `age` lead — the reliance ordering the Basic
forest learned, re-derived once by the pipeline and presented here without any retraining.
Because this document is a downstream **target**, re-rendering it is governed by input hashes:
identical inputs ⇒ an all-skip rebuild, the signature of reproducibility.
