---
title: "Step 04 — The reproducibility capstone"
subtitle: "A targets DAG reloads the Basic model and re-derives its explanation; a re-run skips everything"
params:
  solved: true
format:
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
---

<!-- params$solved unused: step 04 is DEMO-only (no blanks). The flag stays for uniformity with hands-on steps. -->

## Where we are

This is the capstone. We have reloaded, explained, deep-learned, and extracted; now we make
the whole story **reproducible by construction**. A [`targets`](https://docs.ropensci.org/targets/)
pipeline declares the dependencies as a **DAG**: the cohort CSV and the saved model file are
inputs, `load_cohort()` and `reload_model()` turn them into objects, and a single **bridge**
target — `explain_model()` — consumes the reloaded Basic random forest and re-derives its
permutation-VIMP explanation. Anyone who clones this folder runs `tar_make()` and gets the
same explanation, deterministically.

The payoff is what happens on the **second** run: `targets` hashes every input, sees nothing
changed, and **skips all five targets**. That skip is **reproducibility, not fakery** — it is
the opposite of pre-baking a number. Contrast it with the one labeled shortcut in this
workshop, the option-B loss curve we showed as "trained earlier": that was honesty about a
result we did *not* recompute live. Here we *do* recompute on demand, and the skip only means
the inputs are byte-for-byte identical. That distinction — hash-checked invariance vs. a
shown-but-stale number — is the whole point.

<!--
Nota docente (IT): capstone (~20 min, DEMO). Messaggi chiave, nell'ordine:

1. LEGGERE IL DAG (tar_visnetwork): il grafo si legge a colpo d'occhio — cohort_file -> cohort,
   model_file -> model, {model + cohort} -> explanation. Il target `explanation` e' il BRIDGE:
   2 dipendenze a monte (model, cohort), 0 a valle. Formativa min-138: "qual e' il target che
   consuma il modello Basic ed emette la spiegazione?" -> explanation.

2. IL BRIDGE CONSUMA DAVVERO IL MODELLO: explain_model(model, cohort) NON e' decorativo —
   ricalcola la VIMP a permutazione del random forest Basic ricaricato (stesso pattern di step 01,
   stesso trucco del pred_wrapper che instrada attraverso il workflow catturato). Il valore del
   target e' la tibble `vi`, un oggetto pulito e serializzabile (non un ggplot).

3. IL 2o tar_make SALTA TUTTO PERCHE' GLI HASH DEGLI INPUT non sono cambiati (non per via
   dell'output). Questa e' RIPRODUCIBILITA': stesso input -> stesso risultato, verificato per hash.
   E' concettualmente diversa dal mostrare un numero pre-cotto come live (onesta', richiamo min-92:
   l'opzione-B "l'ho allenato prima su GPU"). targets ricalcola SE serve; semplicemente qui non serve.

4. PROVA VISIBILE: `message: false` sopprime i messaggi "skipped" di tar_make, quindi NON ci si fida
   del messaggio. Si prova col VALORE DI RITORNO: tar_outdated() == character(0) (niente da ricostruire)
   + tar_progress() che elenca tutti i target come "skipped". Quello e' stampato, non soppresso.

5. STRETCH (Davide): il device/GPU e' un INPUT CIECO se non e' un target — targets non lo hash-a.
   Cambiare CPU<->GPU non invaliderebbe la pipeline a meno di renderlo esplicito (un target o un dep
   tracciato). Spunto per discutere i limiti dell'invarianza-per-hash.
-->

## Read the DAG {#sec-dag}

`tar_visnetwork()` draws the pipeline graph straight from `_targets.R`: two input **files**
flow into two reloaded **objects**, and both objects flow into the single **explanation**
bridge. Read it left to right — `cohort_file → cohort`, `model_file → model`, and
`{model, cohort} → explanation`. One upstream dependency feeding one downstream consumer:
that shape *is* the reproducibility contract.

```{r dag}
#| code-fold: false
library(targets)

# Read the DAG: input files -> reloaded objects -> the explanation bridge ----
tar_visnetwork()
```

## First build — every target runs {#sec-build}

The first `tar_make()` has nothing cached, so it runs all five targets: it loads the cohort,
reloads the Basic random forest, and computes the permutation VIMP. The bridge target stores
the importance **tibble** (a clean, serializable value — not a plot), which we read back and
plot here.

```{r build}
#| code-fold: false
# First build — every target runs (nothing is cached yet) ----
tar_make()

# The bridge's value: the permutation VIMP of the reloaded Basic RF ----
explanation <- tar_read(explanation)
explanation

# A picture of the same tibble (the target stays a plain object; we plot it here) ----
library(ggplot2)
explanation |>
  dplyr::mutate(Variable = forcats::fct_reorder(Variable, Importance)) |>
  ggplot(aes(x = Importance, y = Variable)) +
  geom_col() +
  labs(
    title = "Permutation VIMP of the reloaded Basic random forest",
    x     = "Drop in ROC-AUC when permuted",
    y     = NULL,
  )
```

`ejection_fraction`, `serum_creatinine`, and `age` lead — exactly the reliance ordering the
Basic forest learned, re-derived here from the reloaded model with no retraining.

## Re-run — nothing is outdated, so the build skips everything {#sec-skip}

Now the proof. `tar_outdated()` returns the names of targets that would need rebuilding;
right after a successful build it returns `character(0)` — **nothing is stale**. We then call
`tar_make()` a second time. Its "skipped" console messages are suppressed (`message: false`),
so we do **not** trust the message: we trust the **return values**. `tar_progress()` reports
the latest run's per-target status, and every target reads `skipped`.

```{r skip}
#| code-fold: false
# tar_outdated() == character(0) means NOTHING needs rebuilding (hash-checked) ----
tar_outdated()

# Second build — messages suppressed; it skips all five targets ----
tar_make()

# Visible proof of the all-skip: every target's status is "skipped" ----
tar_progress()

# And, after the second build, still nothing is outdated ----
tar_outdated()
```

`tar_outdated()` is empty and `tar_progress()` shows all five targets `skipped`. The pipeline
did not recompute the VIMP — not because we hid a stale number, but because the **input hashes
are unchanged**, so the cached result is provably the same one we would get by recomputing.
That is reproducibility: a re-run is all-skip *because nothing changed*, conceptually distinct
from displaying a pre-baked figure as if it were live. Change the CSV by one byte, and
`tar_outdated()` would name `cohort` (and everything downstream) — and only then would
`tar_make()` rebuild.
