---
title: "Step 01 — Import & wrangle"
subtitle: "From 33 raw columns to a tidy modelling tibble"
params:
  solved: true
format:
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
---

## Where we are

Cumulative snapshot: we reload the raw trial data (Step 00), then wrangle it into
the modelling tibble we will use for the rest of the workshop.

<!--
Nota docente (IT): step HANDS-ON (~32 min). Il primo chunk (load) è già visto allo
step 00, qui è "given". Il wrangle è il "Your turn" di QUESTO step: in modalità
blank (`params$solved == FALSE`) si mostra solo lo scheletro `R/yourturn-01.R`
con i `___` da completare dal vivo; in modalità solved gira la soluzione.
Predittori (decisione del controller): age, risk, gender, sod, rx, type.
`site` è stato escluso. L'evento è `1_yes` (~13%).
-->

# Setup — load the raw data {#sec-load}

Given from Step 00.

```{r setup}
#| code-fold: false
library(here)
library(rio)
library(tidyverse)
library(janitor)

indo_raw <- import(here("data-raw", "indo_rct.csv"), setclass = "tibble")
dim(indo_raw)
```

# Your turn — wrangle into a modelling tibble {#sec-wrangle}

Build `indo` from `indo_raw`: clean the names, keep the **outcome** plus the six
predictors (`age`, `risk`, `gender`, `sod`, `rx`, `type`), drop rows with a
missing `age`, and make `outcome` an explicit two-level factor with the **no
event first** (`"0_no"`, `"1_yes"`). The result should have 7 columns.

```{r wrangle-solved, eval=params$solved}
#| code-fold: false
# Wrangle ----
indo <- indo_raw |>
  clean_names() |>
  select(outcome, age, risk, gender, sod, rx, type) |>
  filter(!is.na(age)) |>
  mutate(outcome = factor(outcome, levels = c("0_no", "1_yes")))

glimpse(indo)
```

```{r wrangle-blank, eval=!params$solved, results="asis", echo=FALSE}
# BLANK mode: echo the `___` skeleton from R/yourturn-01.R as a (non-executed)
# code block. We emit it verbatim instead of running it as an R chunk, because
# `___` is not parseable R and knitr parses every R chunk even when eval=FALSE.
cat("```r", readLines(here("R", "yourturn-01.R")), "```", sep = "\n")
```

```{r check, eval=params$solved}
#| code-fold: false
# Quick checks: 7 columns, a 2-level factor, ~13% events.
ncol(indo)
levels(indo$outcome)
count(indo, outcome)
```

We now have a clean, 7-column tibble: `outcome` plus the six predictors. The
event (`1_yes`) is the minority class — roughly 13% — which will steer our metric
choice in the EDA step.
