---
title: "Step 01 — Import & wrangle"
subtitle: "From 13 raw columns to a tidy modelling tibble"
params:
  solved: true
format:
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
---

## Where we are

Cumulative snapshot: we reload the raw clinical data (Step 00), then wrangle it
into the modelling tibble we will use for the rest of the workshop.

<!--
Nota docente (IT): step HANDS-ON (~32 min). Il primo chunk (load) è già visto allo
step 00, qui è "given". Il wrangle è il "Your turn" di QUESTO step: in modalità
blank (`params$solved == FALSE`) si mostra solo lo scheletro `R/yourturn-01.R`
con i `___` da completare dal vivo; in modalità solved gira la soluzione.
Due decisioni chiave: (1) DROPPIAMO `time` (giorni di follow-up) perché è LEAKAGE
— conoscere quanto a lungo uno è stato seguito anticipa di fatto chi muore; (2)
convenzione EVENT-FIRST: il livello "died" è il PRIMO, così tutte le metriche
(roc_auc, pr_auc) trattano la morte come classe positiva di default, senza dover
passare `event_level` ovunque. Evento `died` ~32%.
-->

# Setup — load the raw data {#sec-load}

Given from Step 00.

```{r setup}
#| code-fold: false
library(here)
library(rio)
library(tidyverse)
library(janitor)

hf_raw <- import(here("data-raw", "heart_failure.csv"), setclass = "tibble")
dim(hf_raw)
```

# Your turn — wrangle into a modelling tibble {#sec-wrangle}

Build `hf` from `hf_raw`: clean the names; **drop `time`** — the follow-up
duration *leaks* the outcome (longer follow-up is mechanically tied to who dies);
turn `death_event` into a two-level factor `outcome` with the **event first**
(`"died"`, `"survived"`) and drop the original column; and make the 0/1 clinical
flags explicit factors. The result should have 12 columns (11 predictors +
`outcome`).

```{r wrangle-solved, eval=params$solved}
#| code-fold: false
# Wrangle ----
hf <- hf_raw |>
  clean_names() |>
  select(-time) |>
  mutate(
    outcome = factor(
      if_else(death_event == 1, "died", "survived"),
      levels = c("died", "survived"),
    ),
  ) |>
  select(-death_event) |>
  mutate(across(c(anaemia, diabetes, high_blood_pressure, sex, smoking), factor))

glimpse(hf)
```

```{r wrangle-blank, eval=!params$solved, results="asis", echo=FALSE}
# BLANK mode: echo the `___` skeleton from R/yourturn-01.R as a (non-executed)
# code block. We emit it verbatim instead of running it as an R chunk, because
# `___` is not parseable R and knitr parses every R chunk even when eval=FALSE.
cat("```r", readLines(here("R", "yourturn-01.R")), "```", sep = "\n")
```

```{r check, eval=params$solved}
#| code-fold: false
# Quick checks: 12 columns, a 2-level factor (event first), ~32% events.
ncol(hf)
levels(hf$outcome)
count(hf, outcome)
```

We now have a clean, 12-column tibble: `outcome` plus the eleven predictors. The
event (`died`) is the minority class — roughly 32% — and putting it *first* makes
every downstream metric score it as the positive class automatically. That
minority share is exactly what will steer our metric choice (ROC-AUC **and**
PR-AUC) in the next steps.
