---
title: "Step 02 — Deep learning, honestly"
subtitle: "A live brulee MLP + agnostic SHAP; written-not-trained CNN/RNN/fused nets"
params:
  solved: true
format:
  html:
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
---

## Where we are

We have explained a forest. Now we train a **neural network** on the same cohort — and we
do it **honestly**. The MLP below is trained **live**, small enough to fit in seconds, then
explained with the *exact same* SHAP call we used in step 01: only the model object changes.
That is the whole agnosticism payoff carried forward to a model with no coefficients at all.

The richer architectures — a **1D-CNN** over a signal, an **RNN** over a vitals sequence,
and a **fused** three-branch net that concatenates all three — are **written, not trained**.
We validate their wiring with a single `forward()` **shape-check** on dummy tensors: if the
output is `[5, 2]` (five patients, two logits), the plumbing is correct, and that is all a
shape-check claims. We never pretend to have trained them here.

There is exactly **one** labeled exception: a pre-computed fused-net loss curve, trained
earlier on a GPU and shown with that label attached. And there is a **GPU-vs-CPU** demo
authored write-only — it runs on the classroom GPU machine, not on this CPU box.

<!--
Nota docente (IT): step HANDS-ON + DEMO (~67 min). È il cuore "deep" del workshop.
Sei messaggi chiave, nell'ordine in cui li giochi:

1. HOTSPOT iniziale (~min 60): mostra il chunk `mlp-live`. Le MANOPOLE vengono PRIMA del
   train, non dopo: epochs / learn_rate / hidden_units / penalty. Formativa: con questi
   default il MLP tende a OVERFITTARE -> chiedi "come lo curi?" -> risposta: penalty su,
   hidden giù (o epochs giù). Fallo dal vivo cambiando un solo numero.

2. SHAP RIUSATO (~min 66): la riga `kernelshap(mlp_fit, ...)` è LETTERALMENTE la stessa
   forma del blocco 01 — cambia solo l'oggetto-modello (ora una rete neurale). È la prova
   dell'agnosticismo: nessun coefficiente, stessa interfaccia. Non riscriverla, ricicla.

3. WRITE-NOT-TRAIN (~min 80, Parsons): CNN/RNN/fused sono SCRITTE, non addestrate. L'unico
   "test" che ci concediamo è lo shape-check su tensori finti: se esce `[5, 2]`, il
   cablaggio è giusto. NON addestriamo mai queste reti qui. Il Parsons fa riordinare il
   forward fuso (3 rami -> concat dim=2 -> head).

4. REGOLA DI ONESTÀ (~min 92): l'opzione-B (curva di loss pre-calcolata su GPU) è l'UNICA
   eccezione etichettata del workshop. Tutto il resto o è live (MLP) o è dichiarato
   write-only (CNN/RNN/fused, demo GPU). L'epoca CPU "uccisa" nella demo GPU ERA live:
   parte sul serio, arranca, la fermi. Non è simulata — è una vera epoca CPU interrotta.

5. GPU PAYOFF: la GPU non serve sul MLP minuscolo (gira in secondi su CPU). Serve sul
   carico CNN/fuso, dove il guadagno si vede. Dillo esplicitamente: non vendere la GPU
   sul giocattolo, vendila sul carico vero.

6. torch è già pre-riscaldato nello step 00 (`torch_tensor(1)`); qui il primo tensore non
   paga l'attesa di init. `cuda_is_available()` è FALSE su questa macchina (build CPU):
   in aula, sulla macchina di insegnamento NVIDIA, è TRUE.
-->

## Given — cohort, recipe, SHAP background {#sec-given}

Same cohort wrangle as every step (`clean_names()` first, drop `time`, event-first
`outcome`). We build a recipe the MLP can consume — dummies, zero-variance drop, and
**normalisation** (a neural net needs scaled inputs) — plus the same SHAP background and
predict adapter as step 01.

```{r given}
#| code-fold: true
library(here)
library(rio)
library(tidyverse)
library(janitor)
library(tidymodels)
library(kernelshap)
library(shapviz)
library(torch)
set.seed(123)

hf_tbl <- import(here("data-raw", "heart_failure.csv"), setclass = "tibble") |>
  clean_names() |>
  select(-time) |>
  mutate(
    outcome = factor(
      if_else(death_event == 1, "died", "survived"),
      levels = c("died", "survived"),
    ),
  ) |>
  select(-death_event) |>
  mutate(across(c(anaemia, diabetes, high_blood_pressure, sex, smoking), factor))

data_split <- initial_split(hf_tbl, prop = 0.75, strata = outcome)
train <- training(data_split)

# Recipe + the same SHAP background and adapter as step 01 ----
rec <- recipe(outcome ~ ., data = train) |>
  step_dummy(all_nominal_predictors()) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())
bg <- train |> select(-outcome)
pred_fun <- function(object, X) predict(object, X, type = "prob")$.pred_died
```

## A live MLP — knobs first, then train {#sec-mlp}

The **parameter slide comes before the train**, not after. Four knobs steer a `brulee` MLP:
`epochs` (how long), `learn_rate` (how big each step), `hidden_units` (capacity), and
`penalty` (regularisation). With generous defaults a small net **overfits** this cohort —
the cure is the same one every time: turn `penalty` **up** and `hidden_units` **down** (or
cut `epochs`). We keep it small so it trains live in seconds.

```{r mlp-live}
#| code-fold: false
library(brulee)
set.seed(123)
# Param slide first, then train SMALL (live) ----
mlp_spec <- mlp(hidden_units = 16, epochs = 30, penalty = 0.01, learn_rate = 0.05) |>
  set_engine("brulee") |>
  set_mode("classification")
mlp_fit <- fit(workflow() |> add_recipe(rec) |> add_model(mlp_spec), train)
augment(mlp_fit, testing(data_split)) |> roc_auc(truth = outcome, .pred_died)
```

That `roc_auc` is a genuinely trained number, computed live on held-out patients. Small nets
rarely win on small tabular cohorts, and that honesty is the point: we report what the model
actually does. Note this is a **single held-out estimate**, not a cross-validated one — so
reading it as "the net beats the validated forest" would be apples-to-oranges; the forest's
~0.80 came from proper cross-validation, while this number reflects a single random split.

## The same explainer, a new model {#sec-shap}

Here is the agnosticism payoff again. The MLP has no coefficients to read, but the explainer
does not care — it needs only a `pred_fun` and a background. The line below is the **same
shape** as the SHAP call in step 01; **only the model object changed** (now a neural net).

```{r mlp-shap}
#| code-fold: false
# The SAME agnostic explainer line as block 01 — only the model changed (now a neural net).
set.seed(3)
ks_mlp <- kernelshap(mlp_fit, X = bg[1, ], bg_X = bg, pred_fun = pred_fun)
shapviz(ks_mlp) |> sv_waterfall()
```

The waterfall reads exactly as the forest's did in step 01 — a per-feature decomposition of
one prediction — but now it explains a neural network. One interface, any model behind it.

## Written, not trained — the shape-check {#sec-shapecheck}

The next architectures are **written, not trained**. `R/nn-modules.R` defines three torch
modules — a 1D-CNN branch over a signal, an LSTM branch over a vitals sequence, and a
`fused_net` that runs all three and concatenates them (`dim = 2`) into a two-logit head. We
validate **only the wiring**: feed dummy tensors through `forward()` and check the output
shape. A correct `[5, 2]` proves the widths line up; it claims nothing about accuracy,
because nothing was trained.

```{r shape-check}
#| code-fold: false
source(here("R", "nn-modules.R"))
net   <- fused_net(n_tab = 11, sig_ch = 8, seq_in = 4, hidden = 16)
x_tab <- torch_randn(5, 11)
x_sig <- torch_randn(5, 1, 50)
x_seq <- torch_randn(5, 12, 4)
out   <- net(x_tab, x_sig, x_seq)
out$shape   # expect [5, 2] — five patients, two logits — WITHOUT training
```

Five rows in, a `[5, 2]` tensor out: three branches of different shapes, fused into one
two-class head, and the plumbing holds. That is the entire claim of a shape-check.

## The one labeled exception — option B {#sec-optionb}

Everything above is either live (the MLP) or shape-checked write-only (the nets). The single
exception is this loss curve: it was trained **earlier, on a GPU**, and is replayed from a
pre-computed artifact. It is labeled as such in the plot title, and it is the **only**
pre-baked result in the workshop.

```{r option-b}
#| code-fold: false
# OPTION B — the ONE labeled exception: "I trained this earlier on GPU." NOT live.
readRDS(here("assets", "optionB-loss.rds")) |>
  ggplot2::ggplot(ggplot2::aes(epoch, loss)) +
  ggplot2::geom_line() +
  ggplot2::theme_minimal() +
  ggplot2::labs(title = "Fused-net training loss — pre-computed on GPU (labeled, not live)")
```

## GPU vs CPU — authored write-only {#sec-gpu}

The block below is **authored write-only**: it does not run here (`#| eval: false`). The
GPU payoff is **not** on the tiny MLP — that trains in seconds on a CPU — it is on the CNN
and fused workloads, where the per-epoch cost actually bites. On the classroom NVIDIA/CUDA
machine `torch::cuda_is_available()` returns `TRUE`; on this build box it returns `FALSE`,
so we never execute it here.

The live classroom move: **start on CPU, watch one epoch crawl, kill it, rerun on GPU.**
That killed CPU epoch is a *real* interrupted epoch — not a simulation — and the rerun is
the same code with `cpu = FALSE`. Only the accelerator changes.

```{r gpu-demo}
#| eval: false
#| code-fold: false
# WRITE-ONLY: runs on the classroom GPU machine, not on this CPU build box.
library(luz)

# Wrap the fused net as a luz module (loss + optimizer), then fit on a chosen device ----
fitted <- fused_net |>
  setup(
    loss      = nn_cross_entropy_loss(),
    optimizer = optim_adam,
  ) |>
  set_hparams(n_tab = 11, sig_ch = 8, seq_in = 4, hidden = 16) |>
  fit(
    train_dl,
    epochs       = 40,
    accelerator  = accelerator(cpu = FALSE),   # FALSE -> GPU; flip to TRUE to feel the crawl
  )

# In class: start with cpu = TRUE, watch ONE epoch crawl, interrupt it,
# then rerun with cpu = FALSE and watch the same code fly on the GPU.
```

## Your turn — reorder the fused forward {#sec-yourturn}

A Parsons exercise: the fused `forward()` is the same logic you just shape-checked. Reorder
the scrambled lines so the three branches run, their outputs concatenate along `dim = 2`,
and the head maps the `16 + sig_ch + hidden` concatenation to two logits.

```{r yourturn-02-solved, eval=params$solved, echo=params$solved}
#| code-fold: false
# Your turn (solved): the fused forward() in correct order — 3 branches -> concat(dim=2) -> head.
fused_forward <- function(self, x_tab, x_sig, x_seq) {
  t <- self$tab(x_tab)
  c <- self$cnn(x_sig)
  r <- self$rnn(x_seq)
  self$head(torch_cat(list(t, c, r), dim = 2))
}
fused_forward   # (the body matches fused_net's forward in R/nn-modules.R)
```

```{r yourturn-02-blank, eval=!params$solved, results="asis", echo=FALSE}
cat("```r", readLines(here("R", "yourturn-02.R")), "```", sep = "\n")
```

Build the three branch outputs in any order, but the **concat then head** order is fixed:
you cannot fuse what you have not computed, and the head's input width is the sum of the
three branch widths. Run the shape-check above and the wiring proves itself — `[5, 2]`,
without ever training a thing.
